Supplementary Information

### 2016-12-02 Composition_Ecology.Rmd
Edited thru SuppInfo, light edits to main text

* Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
* John W Williams: Co-conceived paper, and co-wrote the paper

@@

Acknowledgements
---------------------
The authors would like to thanks the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form. In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group. The authors than Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript. This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.  Maps were made using data from Natural Earth (http://naturalearth.com).

>>>>

* Simon J Goring: Co-conceived paper, performed analysis, developed figures and co-wrote the paper
* John W Williams: Co-conceived paper and co-wrote the paper

@@

Acknowledgements
---------------------
The authors would like to thank the large number of individuals who have worked to first, undertake the Public Lands System Survey, to bring the original survey data together, to digitize and standardize much of the survey results, and finally, to assist in interpreting and compiling the data in its present form. In particular, we thank David Mladenoff, Charlie Cogbill, Ed Schools, and the PalEON Settlement Vegetation working group. The authors than Andria Dawson and PalEON participants for helpful comments in the drafting of this manuscript. This material is based upon work carried out by the PalEON Project (http://paleonproject.org), under NSF Award Numbers 1241868 and 1065656.  Maps were made using data from Natural Earth (http://naturalearth.com).

### 2016-12-02 SupplementaryInformation.Rmd
Edited thru SuppInfo, light edits to main text

---
title: "Untitled"
author: "Simon Goring"
date: "November 29, 2016"
output: word_document
---

@@

We include the effects of autocorrelation in the t-tests using the `SpatialPack` function `modified.ttest`. Unfortunately, the `modified.ttest` only performs a pairwise t-test.  Pairwise t-tests are inappropriate for these analyses, because, for the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison.  In this case, only the overlapping points would be included, and it would show zero change since climate is unchanged.

@@

# Climate changes:  1895-1924 vs. 1950-1980

If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: T_min increased XXoC between 1895-1924 and 1950-1980, T_max changed by XXoC, while T_diff changed by XXoC

```{r}

**Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns are percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras. Modern and Historical indicate represent range area (100s of km2) occupied by the tree taxon for the pre-settlement or modern era. Taxa are sorted by the percent area lost (Column 3).  In this analysis, all PLS and FIA grid cells were used and a taxon was assumed have been gained/lost for a given grid cell if that grid cell had data in one time period but not the other.

@@

**Supplemental Table 2**:  As Supplemental Table 1, except that in this analysis, all estimates are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Figure 1.

@@

**Supplemental Table 3**:  As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al. 2010, Baker et al. 1985).  See Supplemental Figure 2 for more information.

@@

**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1980 instead of 1990-2014.  See also Supplemental Figures 3, 4.

@@

**Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Table 2.

@@

**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of using 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e. the time of the PLS surveys).  No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate warming, with estimates of 19th century warming at Fort Leavenworth, Kansas of 0.10oC decade-1 in the winter and 0.07oC decade-1 in the summer (Burnette et al. 2010).  Given this and the ~50 year offset between PLS and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35oC for t_max, +0.5oC for t_min, and +0.15oC for t_diff. This would imply that the reported regional increase in T_min between historical to modern periods (3.3^o^C, Fig. 1) should be further increased to 3.8oC, the reported regional decrease in T_max  (-0.8^o^C) should be reduced to -0.45oC), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C). Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for Tmin and Tdiff and too large for Tmax.  The patterns shown in Supplementary Figure 3 are qualitatively unchanged from those shown in Fig. 4.

@@

**Supplemental Figure 3**:  As Figure 4 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.

@@

**Supplemental Figure 4**:  As Figure 5 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.

>>>>

---
title: "Supplementary Information"
author: "Simon Goring"
date: "November 29, 2016"
output: word_document
---

@@

We include the effects of autocorrelation in the t-tests using the `SpatialPack` function `modified.ttest`. Unfortunately, the `modified.ttest` only performs a pairwise t-test.  Pairwise t-tests are inappropriate for these analyses, because, for the $V_{H}C_{H}$ - $V_{M}C_{H}$ comparison, only the overlapping points would be included, and it would show zero change since climate is unchanged.

@@

# Climate changes between 1895-1924 and 1950-1980  (vs. main paper analysis of 1895-1924 and 1990-2014)

If a 1950-1980 climate normal is used instead of 1990-2014 climate normal, then between 1895-1924 and 1950-1980, the regional climate changes are as follows: ($T_{min}$) increased 0.5^o^C between 1895-1924 and 1950-1980, ($T_{max}$) increased by 0.35^o^C, while T_diff changed by -0.15^o^C

# Supplemental Tables
**Supplemental Table 1**:  Gain and loss of the analyzed tree genera since the pre-settlement era. The first three columns (Gain, Loss, Presence) are expressed as percentages relative to the total area occupied by the taxon in the pre-settlement and modern eras. Modern and Historical indicate represent range area (units: 10^2^ km^2^) occupied by the tree taxon for the pre-settlement or modern era. Taxa are sorted by the percent area lost (Column 3).  In this analysis, all PLS and FIA grid cells were used and a taxon was assumed have been gained/lost for a given grid cell if that grid cell had data in one time period but not the other.

@@

**Supplemental Table 2**:  As Supplemental Table 1, except that in this analysis, all estimates are restricted to the subset of 8 km gridcells that had at least one observation in both the PLS and FIA datasets.  For most taxa, the percentages are similar (within a few percentage points) between Supplemental Tables 1 and 2, except for Quercus, which has a substantially lower estimate of percent area lost (ST1: 46% vs. ST2: 25.7%)  See also Supplemental Figure 1.

@@

<!-- COM: scientific names needed in ST3 -->

**Supplemental Table 3**:  As Table 1 in the main text, except that the assessments of compounding (□), confounding (-), and counteracting (○) interactions are based on a historical climate normal that has been adjusted to account for 19th-century temperature trends reported at early meteorological time series from US military forts (Burnette et al. 2010, Baker et al. 1985).  All entries in this table are identical to those in Table 1, indicating that the assessments of interaction strength are insensitive to this adjustment to mid-19th-century climates. See Supplemental Figure 2 for further details about this analysis.

@@

<!-- COM: ST4: scientific names needed -->
<!-- COM: ST4: Correct column names and formatting for climate variables.  e.g. ppt should be PAnn -->
<!-- COM: Correct formatting for variable names in text.  I have tried to do this where I have seen them, but may have missed some variables.   -->
**Supplemental Table 4**: As Table 1 in the main text, except that the compounding (□), confounding (-), and counteracting (○) assessments are based on a modern climate normal for 1950-1980 instead of 1990-2014. The results for ($T_{min}$), ($T_{diff}$), and ($P_{ann}$) are unchanged. For ($T_{max}$), the sign of the climate change flips from negative (cooling) to positive (warming), indicating that the apparent directionality of ($T_{max}$) change is sensitive to choice of modern climate normal.  (Note that directionality of ($T_{min}$) is unchanged, consistent with 20^th^- and 21^st^-century warming).   See also Supplemental Figures 3, 4.

@@

# Supplemental Figures
**Supplemental Figure 1**:  As Figure 3 in the main text, except that all estimates of gains, losses, and presences are restricted to the subset of gridcells that had observations in both the PLS and FIA datasets.  See also Supplemental Table 2.

@@

**Supplemental Figure 2**:  As Figure 4 in the main text, but with a simple adjustment to check the effects of the decision to use PRISM 1895-1924 climate normals instead of climate normals from the middle 19th century (i.e. a climate normal closer to the time of the PLS surveys).  No gridded instrumental temperature reconstructions are available in the upper Midwest prior to 1895, but early meteorological records from US military forts indicate 19th-century warming, with estimates of  warming at Fort Leavenworth, Kansas of 0.10oC decade^-1^ in the winter and 0.07oC decade^-1^ in the summer (Burnette et al. 2010).  Given this and the ~50 year offset between PLS observations and early PRISM data, the historical climate estimates used here may be biased by approximately +0.35^o^C for ($T_{max}$), +0.5^o^C for ($T_{min}$), and +0.15^o^C for t_diff. This would imply that the reported regional increase in ($T_{min}$) between historical to modern periods (3.3^o^C, Fig. 1) should be further increased to 3.8^o^C, the reported regional decrease in ($T_{max}$)  (-0.8^o^C) should be reduced to -0.45^o^C), and slightly larger reduction in $\Delta t_{diff}$ (-4.25^o^C). Hence, using a 1895-2024 climate normal may cause our estimates of historic climate change to be too conservative for ($T_{min}$) and ($T_{diff}$) and too large for ($T_{max}$).  The patterns shown in Supplementary Figure 3 are qualitatively unchanged from those shown in Fig. 4, suggesting that the findings of this paper are largely insensitive to this adjustment to account for 19th-century rises in temperature.

@@

**Supplemental Figure 3**:  As Figure 4 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014. The results for ($T_{min}$) and ($P_{ann}$) are largely unchanged between this analysis and that shown in Fig. 4, indicating that the finding of tree-climate distribution shifts is robust to choice of modern cliamte normal.  However, ($T_{max}$) shows much closer overlap between the PLS and FIA periods, suggesting that the reported difference in tree-climate distributions for ($T_{max}$) is sensitive to choice of modern climate normal.

@@

**Supplemental Figure 4**:  As Figure 5 in the main text, except that the modern climate normal is for 1950-1980 instead of 1990-2014.  Few differences are apparent between this figure and Figure 5, suggesting that the attribution of tree-climate distribution shifts to changes in climate versus changes in vegetation is insensitive to choice of modern climate normal.
